### 2021-06-29 paper.Rmd
working on the paper again, yay Martin

    Anticlustering is a method to partition data sets into groups 
    in such a way that high between-group similarity and high within-group 
    heterogeneity is obtained. In psychological research, anticlustering is for example 
    used to to divide stimulus sets into equivalent sets, to split tests into parallel 
    halfs, or to divide a group of students into heterogeneous classes. Different 
    objective functions can be optimized in anticlustering applications. K-means 
    anticlustering maximizes the error sum of squares, i.e., the sum of squared 
    Euclidean distances from each element to the center of the group to which it 
    belongs. K-means anticlustering thereby directly maximizes the similarity of 
    the mean values of numeric attributes across groups, but it ignores the spread of 
    the data. This paper presents k-plus anticlustering, a straight-forward 
    extension of the k-means objective function that allows to simultaneously 
    maximize similarity with regard to the mean as well as the variance across 
    groups. It is shown that the bicriterion optimization problem reduces to an 
    augmentation of the input data, leaving the original k-means criterion 
    unchanged. Tests on simulated and real data show that similarity with regard to 
    means and variance criteria can usually be achieved to an astonishing degree 
    with almost no cost to either criterion. It is therefore concluded that the 
    extended model should be preferred to classical k-means anticlustering for 
    optimizing between-group similarity. An exception is given for binary data that 
    does not profit from considering equality in variances.

@@

Anticlustering is a method to partition a set of elements into different 
groups with the focus of obtaining high between-group similarity and 
within-group heterogeneity. Anticlustering reverses the logic of its 
better known twin, cluster analysis that seeks homogeneity within 
clusters and heterogeneity between clusters [@rokach2005; 
@steinley2006]. The method of anticlustering is useful in many 
tasks surrounding the work of researchers in psychology, for example, 
when splitting a test into parts of equal difficulty [@gierl2017], when 
assigning students to work groups [@baker2002], or when assigning 
stimuli to experimental conditions [@lahl2006]. Solving such problems 
"by hand" is a tedious and time-consuming task and the quality of 
manually established groups is usually improvable. Fortunately, these 
problems can be formalized mathematically via several anticlustering 
objective functions [e.g., @brusco2019; @spath1986; @baker2002; 
@fernandez2013], and accessible open source software solutions exist 
[@papenberg2020].

Several anticlustering objectives are available to assess between-group 
similarity or within-group heterogeneity. In many cases, both objectives 
are directly related, that is, maximizing heterogeneity within groups 
simultaneously maximizes between-group similarity. Several 
anticlustering objectives are based pairwise dissimilarity ratings; the 
most prominent criterion of this kind is the *diversity*, which is the 
total sum of dissimilarities between any elements within the same group. 
Note that maximizing the diversity is equivalent to minimizing 
the sum of dissimilarities between elements in different clusters, 
directly illustrating the correspondence between within-group 
heterogeneity and between-group similarity. Another important criterion 
based on pairwise dissimilarities is the dispersion, which is the 
minimum dissimilarity between any two items in the same group 
[@fernandez2013; @brusco2019]. Maximizing the dispersion increases the 
within-group heterogeneity by ensuring that any two elements as 
dissimilar as possible from each other. Note that maximizing the 
dispersion no longer simultaneously maximize between-group similarity. 
In this case, these two potential aims of anticlustering diverge.

@@

A different approach to anticlustering directly works with feature based data 
and does not require a transformation into pairwise dissimilarities. Instead, 
the sum of the squared Euclidean distances between elements and the cluster they 
are assigned to is minimized [@brusco2006branch], most prominently using the 
k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion 
simultaneously maximizes between-cluster separation and within-cluster 
homogeneity [@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed k-means 
objective that is, maximizing the sum of distances between elements and cluster 
centroids -- a data set is partitioned into similar groups, thereby reversing 
the logic of classical clustering. They coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean distances 
from each element to the center of the cluster to which it belongs. As 
@spath1986 noted, k-means anticlustering thereby directly minimizes differences 
with regard to the means of the numeric variables between groups. Hence, groups 
only become similar with regard to the mean of the distribution of the numeric 
attributes, which may perform poorly as a measure of overall similarity; the 
spread of the data is not targeted and may differ between sets. @papenberg2020 
showed that maximizing the diversity (based on the pairwise Euclidean distances) 
is better suited to minimize difference with regard to both the mean and the 
variance of the data; it was# Variance objective function with changed order of the arguments 
# (can be called by sapply() etc.)
var_objective <- function(clusters, features) {
  anticlust::variance_objective(features, clusters)
}
 shown that even an entirely random split usually 
resulted in more similar standard deviations as k-means solely focuses on the 
mean. When requiring that groups are overall similar to each other, this focus 
may be misguided as similar means can be obtained even when the underlying 
distributions are different [@anscombe1973].

@@

K-means anticlustering---i.e., maximizing the SSE---minimizes differences 
between the means of the input variables. I now present a rather 
straight-forward adaption of the SSE to capture differences in variance between 
groups. To motivate the new criterion, we first consider a set of 
one-dimensional observations $(x_1, \ldots, x_N)$. The extension to the general 
case of $P$ attributes will be straight forward

@@

Note that a minor problem arises when ignoring the weights during optimizing the 
k-plus objective. Since their computation involves squaring, the squared 
distances $y_{ij}$ have much larger values, on average, than the raw values 
$x_{ij}$ (also see the axes in Figure 1). Therefore, the $SSE_{Var}$ criterion 
naturally receives greater weight when computing the k-plus criterion than the 
standard $SSE$ criterion. This imbalance is unwarranted because optimizing 
similarity with regard to variances is not more important than optimizing 
similarity with regard to means. A simple solution for this issue is to 
standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing 
$SSE_{kplus}$, e.g. via $z$-standardization.

@@

```{r, fig.cap = "The relationship between the two criteria SSE and $SSE_{Var}$ across partitions is shown. The partition represented by the diamond minimizes the difference in variances between groups; the partition represented by the square minimizes the difference in means between groups; the partition represented by the triangular is a good tradeoff, satisfying both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

@@

For this initial example, I created 14 data points from a univariate normal 
distribution ($M = 15$, $SD = 3$) and generated all `r n_partitions(14, 2)` 
possible ways to partition the 14 data points into $K = 2$ equal-sized groups. 
Figure 1 shows the scatterplot of the criteria SSE and $SSE_{Var}$ across 
partitions. Note that for both criteria, a larger value is better, i.e., 
indicates higher similarity with regard to means and variances, respectively. 
Therefore, to obtain overall between-group similarity, a partition should be 
chosen that minimizes the difference with regard to both means and variances, 
i.e., a partition that is depicted in the upper right corner of the scatterplot. 
However, optimal k-means anticlustering would create two groups that are not 
particularly similar with regard to their variance. The partition that is 
represented by the squared symbol has the overall best highest SSE, but is 
outperformed by many other partitions with regard to $SSE_{Var}$. Similarly, the 
partition that maximizes $SSE_{Var}$ (the triangular symbol in Figure 1) is not 
well-suited to optimize SSE at the same time. The diamond symbol represents the 
partition that maximizes $SSE_{kplus}$, i.e., sum of SSE and $SSE_{Var}$. This 
partition is located exactly in the right upper corner and yields `r 
df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum 
SSE, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global 
maximum value of $SSE_{Var}$. Hence, this initial example shows that maximizing 
the combined k-plus criterion may well be suited to optimize both similarity 
with regard to means as well as variances between groups. Later, I will show 
that this result also holds for larger data sets, more groups, and more 
features.

>>>>

    Anticlustering refers to a collection of methods that partition data sets into disjunct groups. The partitioning process may pursue two goals: (a) high between-group similarity and (b) high within-group heterogeneity. Many anticlustering objectives simultaneously pursue both goals while other methods focus on a particular aspect. Recently, Brusco, Cradit and Steinley (2020) presented a highly effective bicriterion optimization algorithm that focuses primarily on maximzing within-group heterogeneity. The present paper introduces a novel bicriterion objective for optimizing between-group similarity. *K-plus anticlustering* is a straightforward extension of k-means anticlustering, which maximizes the sum of squared Euclidean distances between each data point and the center of the cluster to which it belongs. Thereby, k-means anticlustering directly minimizes the difference between the mean values of numeric attributes across clusters, but neglects optimizing similarity with regard to the variance of the data sets. K-plus anticlustering extends the k-means objective by adding a term that aims to simultaneously optimize similarity with regard to the variance. Surprisingly, it is shown that the combined objective (i.e., the bicriterion) can be reduced to an augmentation of the input data, leaving the original k-means criterion unchanged. Tests on simulated and real data show that k-plus anticlustering achieves high similarity both with regard to mean attribute values as well as the variance, usually with no noticable cost to either criterion. It is therefore concluded that the extended model should be preferred to classical k-means anticlustering when optimizing between-group similarity.

@@

Anticlustering is a collection of methods to partition elements into disjunct groups, with the goal of obtaining high between-group similarity or within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Oftentimes, both goals coincide: several anticlustering objectives imply that similarity between groups is maximal whenever similarity within groups is minimal [@brusco2019; @papenberg2020; @feo1990]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and heterogeneity between clusters [@rokach2005; @steinley2006]. Whereas a variety of different approaches exist for cluster analysis such as hierarchical methods, model-based clustering, or network approaches [see @brusco_emergent_2012 for an overview], anticlustering is usually formalized as a partitioning process that divides a set of elements exhaustively into disjunct groups in such a way that an objective function---representing between-group similarity and/or within-group heterogeneity---is optimized [@papenberg2020; @spath1986; @brusco2019].

Anticlustering is useful in many tasks surrounding research work in psychology. For example, anticlustering can be used to split a test into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to experimental conditions [@lahl2006]. For an overview of anticlustering applications in psychology and other areas, see @brusco2019 and @papenberg2020. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual group assignment is usually suboptimal. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

---

**TODO; hier weitermachen**

Several anticlustering objectives are based pairwise dissimilarity ratings; the most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group. Another important criterion based on pairwise dissimilarities is the dispersion, which is the minimum dissimilarity between any two items in the same group [@fernandez2013; @brusco2019]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements as dissimilar as possible from each other.

@@

A different approach to anticlustering directly works with feature based data 
and does not require a transformation into pairwise dissimilarities. Instead, 
the sum of the squared Euclidean distances between elements and the cluster they 
are assigned to is minimized [@brusco2006branch], most prominently using the 
k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion 
simultaneously maximizes between-cluster separation and within-cluster 
homogeneity [@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed k-means 
objective that is, maximizing the sum of distances between elements and cluster 
centroids -- a data set is partitioned into similar groups, thereby reversing 
the logic of classical clustering. They coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean distances 
from each element to the center of the cluster to which it belongs. As 
@spath1986 noted, k-means anticlustering thereby directly minimizes differences 
with regard to the means of the numeric variables between groups. Hence, groups 
only become similar with regard to the mean of the distribution of the numeric 
attributes, which may perform poorly as a measure of overall similarity; the 
spread of the data is not targeted and may differ between sets. @papenberg2020 
showed that maximizing the diversity (based on the pairwise Euclidean distances) 
is better suited to minimize difference with regard to both the mean and the 
variance of the data; it was shown that even an entirely random split usually
resulted in more similar standard deviations as k-means solely focuses on the 
mean. When requiring that groups are overall similar to each other, this focus 
may be misguided as similar means can be obtained even when the underlying 
distributions are different [@anscombe1973].

@@

K-means anticlustering---i.e., maximizing the SSE---minimizes differences 
between the means of the input variables. I now present a rather 
straight-forward adaption of the SSE to capture differences in variance between 
groups. To motivate the new criterion, we first consider a set of 
one-dimensional observations $(x_1, \ldots, x_N)$. The extension to the general 
case of $P$ attributes will be straight forward.

@@

Note that a problem may arise when ignoring the weights during optimizing the
k-plus objective. Since their computation involves squaring, the squared 
distances $y_{ij}$ have much larger values, on average, than the raw values 
$x_{ij}$ (also see the axes in Figure 1). Therefore, the $SSE_{Var}$ criterion 
naturally receives greater weight when computing the k-plus criterion than the 
standard $SSE$ criterion. This imbalance is unwarranted because optimizing 
similarity with regard to variances is not more important than optimizing 
similarity with regard to means. A simple solution for this issue is to 
standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing 
$SSE_{kplus}$, e.g. via $z$-standardization.

@@

```{r, fig.cap = "The relationship between the two criteria SSE and $SSE_{Var}$ across partitions is shown. The partition represented by the triangular minimizes the difference in variances between groups; the partition represented by the square minimizes the difference in means between groups; the partition represented by the diamond is a good tradeoff, satisfying both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

@@

For this motivating example, I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $SSE$ and $SSE_{Var}$ across partitions. Note that for both criteria, a larger value is better, i.e., indicates higher similarity with regard to means and variances, respectively. Therefore, to obtain overall between-group similarity, a partition should be chosen that is depicted in the upper right corner of the scatterplot. However, optimal k-means anticlustering would create two groups that are not particularly similar with regard to their variance. The partition that is represented by the squared symbol has the overall best highest $SSE$, but is outperformed by many other partitions with regard to $SSE_{Var}$. Similarly, the partition that maximizes $SSE_{Var}$ (the triangular symbol in Figure 1) is not well-suited to optimize SSE at the same time. The diamond symbol represents the partition that maximizes $SSE_{kplus}$, i.e., sum of SSE and $SSE_{Var}$. This partition is located in the right upper corner and yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum SSE, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $SSE_{Var}$.

This initial example shows that maximizing the combined k-plus criterion may well be suited to optimize both similarity with regard to means as well as variances between groups. Later, I will show that this result also holds for real data sets, using more groups and a larger number of numeric features.
